### Part a

Fill in the blanks in this piece of text, using words and phrases from Table\ \@ref(tab:Words1Table).
   

   
> * The decision-making process begins with making an assumption about the population __________.  
> * This means we know what to __________ from the sample __________.   
> * We never know exactly what value of the statistic we will see in the sample, because of __________.   
> * But we can have some of idea of what values are reasonable to expect.  
> * Then we take the __________ (that is, we make the observations).  
> * Then we __________ what the sample statistic that we observed... to the sample statistic we expected.  
> * If what we observe is inconsistent with what was expected, then the the assumption is __________ true.  
> * However, if what we observe is __________ with what was expected, then the the assumption is __________ true.



```{r Words1Table, echo=FALSE}
Words1 <- array( "", dim = c(2, 5))

Words1[1, ] <- c("compare", "probably", "expect", "sampling variation", "consistent")
Words1[2, ] <- c("unlikely to be", "sample", "parameter", "statistic", "  ")


if( knitr::is_latex_output() ) {
  kable(Words1,
      format = "latex",
      align = "c",
      booktab = TRUE,
      caption = "Words and phrases to insert.") %>%
    kable_styling(full_width = FALSE,
                font_size = 10)
}
```


### Part b

Fill in the blanks in this piece of text,
using words and phrases from Table\ \@ref(tab:Words3Table).

> * Suppose I wish to see if a six-sided die is fair.  
> * My __________ would be that the proportion of odd numbers is $0.5$.
>   This value is a __________.   
> * The __________ would be "the proportion of odd numbers in the sample".  
> * We would __________ that the value of the statistic would be near $0.5$ (but maybe not exactly $0.5$).  
> * If I roll the die $100$ times and observe **53** odd numbers, this seems __________ with what I would expect if the assumption was true.  
> * If I roll the die $100$ times and observe **zero** odd numbers, this seems __________ with what I would expect if the assumption was true.


```{r Words3Table, echo=FALSE}
Words3 <- array( "", dim = c(2, 3))

Words3[1, ] <- c("assumption", "consistent", "inconsistent")
Words3[2, ] <- c("statistic", "expect", "parameter")


if( knitr::is_latex_output() ) {
  kable(Words3,
      format = "latex",
      booktab = TRUE,
      align = "c",
      caption = "Words and phrases to insert.") %>%
  kable_styling(full_width = FALSE,
                font_size = 10)}
```


