### Part a

Fill in the blanks in this piece of text, using words and phrases from Table\ \@ref(tab:Words1Table).
   

   
> * The decision-making process begins with making an assumption about the population _____.
> * This means we know what to _____ from the sample _____. 
> * We never know exactly what value of the statistic we will see in the sample, because of _____. 
> * But we can have some of idea of what values are reasonable to expect.
> * Then we take the _____ (that is, we make the observations).
> * Then we _____ what the sample statistic that we observed... to the sample statistic we expected.
> * If what we observe is inconsistent with what was expected, then the the assumption is _____ true.
> * However, if what we observe is _____ with what was expected, then the the assumption is _____ true.



```{r Words1Table, echo=FALSE}
Words1 <- array( "", dim=c(3, 3))

Words1[1, ] <- c("compare", "probably", "expect")
Words1[2, ] <- c("unlikely to be", "sample", "parameter")
Words1[3, ] <- c("sampling variation", "statistic", "consistent")


if( knitr::is_latex_output() ) {
  kable(Words1,
      format = "latex",
      align = "c",
      booktab = TRUE,
      caption = "Words and phrases to insert") %>%
  kable_styling(full_width=FALSE)
}
```


### Part b

Locate the correct boxes in the **decision-making process** to place the words and phrases given in Table\ \@ref(tab:Words2Table).



```{r Words2Table, echo=FALSE}
Words2 <- array( "", dim = c(2, 3))

Words2[1, ] <- c("Observation of sample statistics",
                 "Expectation: sample statistic",
                 "Consistent")
Words2[2, ] <- c("Assumption: population parameter",
                 "Yes: Suppports assumption",
                 "No: Suppports assumption")



if( knitr::is_latex_output() ) {
  kable(Words2,
      format = "latex",
      align = "c", 
      booktab = TRUE,
      caption = "Words and phrases to insert into the decision-making process") %>%
  kable_styling(full_width=FALSE)
}
```




```{r DecisionFlowBlanks, echo=FALSE, fig.cap="A way to make decisions", fig.align="center", fig.width=6.5, fig.height=3, out.width="85%"}

par( mar=c(0.15, 0.15, 0.15, 0.15))
openplotmat()

pos <- array(NA, dim = c(6, 2))
pos[1, ] <- c(0.10, 0.85) # Assumption
pos[2, ] <- c(0.40, 0.85) # Expectation
pos[3, ] <- c(0.40, 0.15) # Observation
pos[4, ] <- c(0.40, 0.50) # Consistency?
pos[5, ] <- c(0.80, 0.85) # YES
pos[6, ] <- c(0.80, 0.15) # NO

straightarrow(from = pos[1,], 
              to = pos[2,], 
            #segment = c(0.4, 0.6),
            lty = 1, 
            lwd = 2)
straightarrow(from = pos[4,], 
              to = pos[2,], 
            lcol = "black",
           lty = 2,
            lwd = 2)
straightarrow(from = pos[4,], 
              to = pos[3,], 
#            curve = 0.8,
            arr.pos = 0.5, # Then cover with box
           lty = 2)
#straightarrow(from=pos[4,], to=pos[5,], 
#            lcol=grey(0.4),
##            curve = -0.8,
#            arr.pos=0.5, # Then cover with box
#            #segment = c(0.25, 0.75),
#            lty=2)
curvedarrow(from = pos[4,] + c(0, 0.065), 
            to = pos[5,], 
            curve = 0.35,
            arr.pos = 0.5, # Then cover with box
            #segment = c(0.25, 0.75),
            lty = 2)
curvedarrow(from = pos[4,] - c(0, 0.065), 
            to = pos[6,] , 
            curve = -0.35,
            arr.pos = 0.5, # Then cover with box
            #segment = c(0.25, 0.75),
            lty = 2)
#straightarrow(from=pos[4,], to=pos[6,], 
#            lcol=grey(0.4),
##            curve = -0.8,
#            arr.pos=0.5, # Then cover with box
#            #segment = c(0.25, 0.75),
#            lty=2)

#points( mean(pos[2:3, 1]), 0.23, pch="\u2714", cex=2, col="grey") # CHECKMARK
#points( mean(pos[2:3, 1]), 0.8,  pch="\u2718", cex=2, col="grey") # CROSS
#text( mean(pos[2:3, 1]), 0.29, "If yes", col="grey") # CHECKMARK
#text( mean(pos[2:3, 1]), 0.73,  "If no", col="grey") # CROSS
textrect( pos[4, ],
          radx = 0.11,
          rady = 0.1,
          shadow.size = 0,
          lcol = "snow2",
          box.col = "snow2",
          lab = "Decision",
          col = grey(0)) # CHECKMARK
textrect( pos[5, ], 
          radx = 0.11,
          rady = 0.1,
          shadow.size = 0,
          lcol = "darkseagreen2",
          box.col = "darkseagreen2",
          lab = "(Conclusion A)", 
          col = grey(0.5)) # CROSS
textrect( pos[6, ], 
          radx = 0.11,
          rady = 0.1,
          shadow.size = 0,
          lcol = "darksalmon",
          box.col = "darksalmon",
          lab = "(Conclusion B)", 
          col = grey(0.5)) # CROSS

textrect( pos[1,], 
          lab = "(Step 1)",
          radx = 0.11,
          rady = 0.1,
          shadow.size = 0,
          lcol ="slategray1",
          box.col = "slategray1",
          cex = 1, 
          col = grey(0.5))
textrect( pos[2,], 
          lab = "(Step 2)", 
          radx = 0.11,
          rady = 0.1,
          shadow.size = 0,
          lcol = "slategray2",
          box.col = "slategray2",
           cex = 1, 
          col = grey(0.5))
textrect( pos[3,], 
          box.col = "slategray3",
          lcol = "slategray3",
          shadow.size=0,
          radx = 0.11,
          rady = 0.1,
          lab = "(Step 3)",
          cex = 1, 
          col = grey(0.5))

```




### Part c

Fill in the blanks in this piece of text,
using words and phrases from Table\ \@ref(tab:Words3Table).

> * Suppose I wish to see if a six-sided die is fair.
> * My _____ would be that the proportion of odd numbers is $0.5$. 
>   This value is a ______.
> * The _____ would be "the proportion of odd numbers in the sample".
> * We would _____ that the value of the statistic would be near $0.5$ (but maybe not exactly $0.5$).
> * If I roll the die $100$ times and observe **53** odd numbers, this seems _____ with what I would expect if the assumption was true.
> * If I roll the die $100$ times and observe **zero** odd numbers, this seems _____ with what I would expect if the assumption was true.


```{r Words3Table, echo=FALSE}
Words3 <- array( "", dim = c(2, 3))

Words3[1, ] <- c("assumption", "consistent", "inconsistent")
Words3[2, ] <- c("statistic", "expect", "parameter")


if( knitr::is_latex_output() ) {
  kable(Words3,
      format = "latex",
      booktab = TRUE,
      align = "c",
      caption = "Words and phrases to insert") %>%
  kable_styling(full_width=FALSE)
}
```


